---
title: "Tutorial Part IV - Movement - Solution Exercise 2"
author: "Stephanie Kramer-Schadt & Cédric Scherer"
date: "`r Sys.setlocale('LC_TIME','C'); paste('Last Update', format(Sys.time(), '%B %e, %Y'))`"
output:
  rmdformats::readthedown:
    highlight: kate
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: console
params:
  date: !r Sys.Date()
---

<style>
h1 {
  color: Orange ;
}
h2, h3, h4, h5, h6, legend {
  color: Indigo ;
}
p {
  line-height:170%;
}
#sidebar h2 {
  background-color: Indigo;
}
code {
  color: Indigo ;
}
.exercise {
  color: #824b00;
}
</style>

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      fig.width = 9, fig.height = 6, dev = "ragg_png")
#knitr::opts_knit$set(root.dir = 'C:/Users/kramer/PopDynIZW Dropbox/Steph Kramer')
#or on GUI: knit -> knit directory -> project directory
```

# Exercise 4.2
There is telemetry data available from three animals (DieTiere_32633.shp) in the
folder data-raw. Additionally, land cover (Landnutzung_3035.shp) and rivers (linearegew_32633.shp) are available in the geo-raw folder. 
Land use has a column 'NS1' with coded info on 1 (urban areas), 2 (agriculture), 
3(forest), 4(swamps), 5 (water bodies).

After loading libraries, the data and making a plot, please do the following:
a) Calculate their home ranges.
b) Buffer the home range with the mean displacement per step of all three animals
c) Clip the home ranges with the underlying landscape and calculate the percentage
of land use classes within the home range (hint: use st_intersection)



## Get started

```{r}
# Open libraries
## general basics
library(here)
library(ggplot2)

## the basic packages for spatial analyses and viz
library(sf)
library(terra)
library(tmap)

## for telemetry and home range analyses
library(move)
library(adehabitatHR) 
library(sp)


##
#library(dismo)
#library(raster)
#library(GISTools)
#library(rgdal)
#library(maptools)
#library(rgeos)
#library(rgl)
#library(raster)
#library(rasterVis)
```

### Define workspace and folder structure

```{r}
work_wd <- here()
data_raw_wd <- here('data-raw')
geo_raw_wd  <- here('data-raw/geo-raw')
output_wd   <- here('output')
```

### Load data

```{r}
# if you are not sure about the file names, check the folder content with
# list.files(data_raw_wd)
animals <- st_read(dsn=data_raw_wd,layer="DieTiere_32633")  
landcov <- st_read(layer="Landnutzung_3035", dsn=geo_raw_wd)    
rivers  <- st_read(layer="linearegew_32633", dsn=geo_raw_wd)   
```

### Define or transform CRS

```{r}
lc_32633 <- st_transform(landcov, 32633)
```

### Make a plot of the data

```{r}
#Create a color code for the 5 land use classes
mycol <- c('red','antiquewhite','chartreuse4','cornsilk4','blue')

plot(lc_32633[,'NS1'], col=mycol[lc_32633$NS1],border='transparent')
plot(animals[,1],col=animals$ID,add=TRUE) 
plot(rivers[,1], col= 'blue', lwd=0.1,add=T)

```

```{r}
#Create a color code for the 5 land use classes
mycol <- c('red','antiquewhite','chartreuse4','cornsilk4','blue')

# TODO - plot mit colorcode mycol funktioniert nicht....

ggplot(lc_32633) + 
  geom_sf(data=lc_32633[,'NS1'], aes(fill  = mycol[NS1]), color='transparent') +
  geom_sf(data=animals) + 
  geom_sf(data=rivers, color='blue') +
  theme_classic()
```

```{r}
#Create a color code for the 5 land use classes
mycol <- c('red','antiquewhite','chartreuse4','cornsilk4','blue')
myid  <- c('yellow','pink','red')

tmap_mode(mode = "view")
tm_shape(shp = lc_32633[,'NS1']) + 
  tm_polygons(col = "NS1", palette = mycol) +
tm_shape(shp = animals[,'ID']) + 
  tm_dots(col = "ID", palette = myid) +
tm_shape(shp = rivers[,1]) + 
  tm_lines(col = 'blue') 
```


# a) & b)
First , we retrieve the steplength. For this, transform the sf-Object into a move-Object.

```{r}
anim_coords <- as.data.frame(st_coordinates(animals)) #returns a list
animals$X <- anim_coords$X
animals$Y <- anim_coords$Y
animals_ng <- st_drop_geometry(animals)

animals_move <- move(x=animals_ng$X, y=animals_ng$Y,
                     time = as.POSIXct(animals_ng$MYDATUM, 
                                       format = "%Y-%m-%d 8:0:0", tz="CET"),
                     proj=CRS("+init=epsg:32633"),
                     data=animals_ng, animal = animals_ng$ID, sensor='GPS')
plot(animals_move, type='l') 

# since we have 3 animals, a list with 3 elements is returned
steplength_per_animal <- distance(animals_move)
mean_stepl_p_anim <- lapply(steplength_per_animal,mean)
(mean(unlist(mean_stepl_p_anim)))
```

The mean step length is approx. 300 m.

```{r}
# for using mcp() in {adehabitatHR}, convert to SpatialPointsDataframe of {sp}
animals_sp <-as(animals, "Spatial")
mcp <- mcp(animals_sp[,'ID'], percent=95) 

# the buffer command needs however an sf object
mcp_sf  <- as(mcp, "sf")
mcp_buf <- st_buffer(mcp_sf, 300)

# the plot is correct
plot(mcp)                         
plot(animals_sp, add=TRUE, col= animals_sp$ID) 
plot(mcp_buf[,1], add=TRUE)
plot(animals_sp, add=TRUE, col= animals_sp$ID) 
plot(mcp, add=TRUE)

# save buffer
st_write(obj = mcp_buf, 
         dsn = output_wd, 
         layer = 'mcp_buffer_32633',
         driver = 'ESRI Shapefile', 
         delete_layer = TRUE)
```

c) clip the home ranges with the land use

```{r}
# check the total area of the buffer in km2:
buf_area <- st_area(mcp_buf)/1e6

# clip
mcp_buf_lc <- st_intersection(lc_32633, mcp_buf) 
# have a look at the object:
mcp_buf_lc

# now, for each animal id, summarize the area per NS1-category:
tapply(mcp_buf_lc$AREA,mcp_buf_lc$id,sum)


```



# END

---

<details><summary>Session Info</summary>

```{r sessionInfo}
Sys.time()
sessionInfo()
```

</details>
