---
title: "Tutorial Part IV - Movement - Solution Exercise 2"
author: "Stephanie Kramer-Schadt & Cédric Scherer"
date: "`r Sys.setlocale('LC_TIME','C'); paste('Last Update', format(Sys.time(), '%B %e, %Y'))`"
output:
  rmdformats::readthedown:
    highlight: kate
    toc_depth: 2
    toc_float: true
editor_options:
  chunk_output_type: console
params:
  date: !r Sys.Date()
---

<style>
h1 {
  color: Orange ;
}
h2, h3, h4, h5, h6, legend {
  color: Indigo ;
}
p {
  line-height:170%;
}
#sidebar h2 {
  background-color: Indigo;
}
code {
  color: Indigo ;
}
.exercise {
  color: #824b00;
}
</style>

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, 
                      fig.width = 9, fig.height = 6, dev = "ragg_png")
#knitr::opts_knit$set(root.dir = 'C:/Users/kramer/PopDynIZW Dropbox/Steph Kramer')
#or on GUI: knit -> knit directory -> project directory
```

# Exercise 4.2
There is telemetry data available from three animals (DieTiere_32633.shp) in the
folder data-raw. Additionally, land cover (Landnutzung_3035.shp) and rivers (linearegew_32633.shp) are available in the geo-raw folder. 
Land use has a column 'NS1' with coded info on 1 (urban areas), 2 (), 3(),
4(), 5 (water bodies).

After loading libraries, the data and making a plot, please do the following:
a) Calculate their home ranges and the centroid of the home range (hint: st_centroid)
b) Buffer the home range with the mean displacement per step of all three animals
c) Clip the home ranges with the underlying landscape and calculate the percentage
of land use classes within the home range (hint: use st_intersection)


## Get started

```{r}
# Open libraries
## general basics
library(here)
library(ggplot2)

## the basic packages for spatial analyses and viz
library(sf)
library(terra)
library(tmap)

## for telemetry analyses
library(move)
library(adehabitatHR)         

##
library(sp)
library(dismo)
library(raster)
library(GISTools)
library(rgdal)
library(maptools)
library(rgeos)
library(rgl)
library(raster)
library(rasterVis)
```

### Define workspace and folder structure

```{r}
work_wd <- here()
data_raw_wd <- here('data-raw')
geo_raw_wd  <- here('data-raw/geo-raw')
output_wd   <- here('output')
```

### Load data

```{r}
# if you are not sure about the file names, check the folder content with
# list.files(data_raw_wd)
animals <- st_read(dsn=data_raw_wd,layer="DieTiere_32633")  
landcov <- st_read(layer="Landnutzung_3035", dsn=geo_raw_wd)    
rivers  <- st_read(layer="linearegew_32633", dsn=geo_raw_wd)   
```

### Define or transform CRS

```{r}
lc_32633 <- st_transform(landcov, 32633)
```

### Make a plot of the data

```{r}
#Create a color code for the 5 land use classes
mycol <- c('red','antiquewhite','chartreuse4','cornsilk4','blue')

# TODO # immer scheitere ich an den plots!!!!!

plot(lc_32633[,'NS1'], col=mycol[lc_32633$NS1],border='transparent')
plot(animals[,1],col=animals$ID,add=TRUE) 
plot(rivers[,1], col= 'blue', lwd=0.1,add=T)

```

```{r}
#Create a color code for the 5 land use classes
mycol <- c('red','antiquewhite','chartreuse4','cornsilk4','blue')

# TODO - plot mit colorcode mycol funktioniert nicht....

ggplot(lc_32633) + 
  geom_sf(data=lc_32633[,'NS1'], aes(fill  = mycol[NS1]), color='transparent') +
  geom_sf(data=animals) + 
  geom_sf(data=rivers, color='blue') +
  theme_classic()
```


```{r}
#Create a color code for the 5 land use classes
mycol <- c('red','antiquewhite','chartreuse4','cornsilk4','blue')
myid  <- c('yellow','pink','red')

tmap_mode(mode = "view")
tm_shape(shp = lc_32633[,'NS1']) + 
  tm_polygons(col = "NS1", palette = mycol) +
tm_shape(shp = animals[,'ID']) + 
  tm_dots(col = "ID", palette = myid) +
tm_shape(shp = rivers[,1]) + 
  tm_lines(col = 'blue') 
```



# END

---

<details><summary>Session Info</summary>

```{r sessionInfo}
Sys.time()
sessionInfo()
```

</details>
