---
title: "Tutorial Part IV — Getting started with animal movement"
author: "Stephanie Kramer-Schadt"
date: "`r Sys.setlocale('LC_TIME','C'); paste('Last Update', format(Sys.time(), '%B %e, %Y')) `" 
        #"`r Sys.Date()`" # 
output:
  rmdformats::readthedown:
    highlight: kate
    code_folding: show
    toc_depth: 4
    toc_float: true
editor_options:
  chunk_output_type: console
params:
  date: !r Sys.Date()
---

<style>
h1 {
  color: Orange ;
}
h2, h3, h4, h5, h6, legend {
  color: Indigo ;
}

#sidebar h2 {
  background-color: Indigo;
}
</style>
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

In this tutorial, I will show you how to get started with (real) animal movement data. 
To put it simple, animal movement data are just points in time and space, i.e. re-locations
of an individual can be provided as a *data.frame* with four columns. An
identifier for the individual is only needed if you have different animals in one 
data set, i.e. if you have collared several individuals with the same collar or
already appended several data sets:

Identifier    | Timestamp           | x_coordinate    | y_coordinate    
------------- | ------------------- | --------------- | ---------------   
Fox_1         | 2022-01-01 08:10:04 | 13.47027        | 52.497802
Fox_1         | 2022-01-01 08:15:04 | 13.47015        | 52.497813


From [Course2_SpatialR](https://github.com/stephkramer/Course2_SpatialR), 
you might remember that these can be imported 
as simple .txt or .csv files. Once you know in which coordinate reference system (CRS)
your coordinates are stemming from, you can **assign** it to the *data.frame*, thereby
you are creating a geo-referenced data set: a *SpatialPointsDataframe* - Object (if
you work with the R-package `{sp}`) or an *sf* - Object (with package `{sf}`). 
Looking at these x- and y-coordinates, you might remember that they look like 
angular units, e.g. decimal degrees, and the x-coordinate refers to longitude and the
y-coordinate to latitude. Hence, we are dealing with EPSG-code 4326. 
Once you have imported and georeferenced your data, you can plot and thoroughly check the data.

Checking your data is a mandatory prerequisite before any analysis (see 
Zuur et al. 2010). With your movement data, you might have outliers in space 
(e.g. when you test your collars in location A which might be hundreds of kilometers from your actual trapping site), your data might not have been regularly sampled because the animal was hiding and
there was no access to the satellite (weak GPS-signal) etc etc. This uneven 
sampling can affect calculations of speed, for example, and therefore a series of packages
have been developed to deal with those issues in movement data, like `{move}`. 

The most prominent movement analyses comprise home range estimation, calculations of
habitat preference, behavioral analyses (e.g. Hertel et al. 2021) and
detection of movement syndromes (personality differences) (Michelangeli et al. 2021).
Some of the packages listed in the next chapter are designed for these analyses. 

```{r,eval=FALSE,echo=FALSE,results='hide'}

[TODO later - exchange Q with Caro's foxes for habitat selection, ref PhD C Scholz]

```


In the following, we will step by step start with loading and exploring movement data
of female red fox (*Vulpes vulpes*) 'Q von Stralau', who had a small home range 
in Berlin, Germany. She was collared in January 2018 and had a very stable daily routine
as shown by 4 (20) min relocation intervals. Her location data are stored 
in the file with the tag number 'tag5334_gps.txt'. For any details on the data please refer to Kimmig (2021) and Scholz (2020).

We will, however, only use the data of one month in this exercise, 
as data sets quickly get too big. Please note: the courses
[Course1_IntroR](https://github.com/stephkramer/Course1_IntroR) and [Course2_SpatialR](https://github.com/stephkramer/Course2_SpatialR) are 
obligatory for this tutorial.

 
## Useful (web)sites and reading

*  For analysis of telemetry data: packages **adehabitatHR, adehabitatLT, move,recurse, momentuHMM, moveHMM, ctmm, amt**
<br>

**Methods papers** <br>

* Joo, R, Boone, ME, Clay, TA, Patrick, SC, Clusella-Trullas, S, Basille, M. Navigating through the r packages for movement. J Anim Ecol. 2020; 89: 248– 267. https://doi.org/10.1111/1365-2656.13116
* Zuur, A.F., Ieno, E.N. and Elphick, C.S. (2010), A protocol for data exploration to avoid common statistical problems. Methods in Ecology and Evolution, 1: 3-14. https://doi.org/10.1111/j.2041-210X.2009.00001.x

**Example papers** <br>

* Hertel, AG, Royauté, R, Zedrosser, A, Mueller, T. Biologging reveals individual variation in behavioural predictability in the wild. J Anim Ecol. 2021; 90: 723– 737. https://doi.org/10.1111/1365-2656.13406


* Kimmig, S (2021). The ecology of red foxes (*Vulpes vulpes*) in urban environments, PhD thesis, FU Berlin. https://refubium.fu-berlin.de/handle/fub188/32478

*  Michelangeli, M., Payne, E., Spiegel, O., Sinn, D. L., Leu, S. T., Gardner, M. G., & Sih, A. (2021). Personality, spatiotemporal ecological variation and resident/explorer movement syndromes in the sleepy lizard. Journal of Animal Ecology, 00, 1– 14. https://doi.org/10.1111/1365-2656.13616

* Scholz, C (2020). The ecology of red foxes (*Vulpes vulpes*) in anthropogenic
landscapes. PhD thesis, FU Berlin. https://refubium.fu-berlin.de/handle/fub188/29984

   

# Getting started

To follow the tutorial, you can either clone or download the repository at
https://github.com/stephkramer/Course4_MoveQ
or you create your own R-project, copy the raw data and type the code chunks into an R-Script.
Please refer to the section on using R-projects in [Course2_RSpatial](https://github.com/stephkramer/Course2_SpatialR).

If you start with your own R-project, I strongly recommend to use the d6-package 
Cedric Scherer provided. This package automatically sets up the ideal folder structure:
https://github.com/EcoDynIZW/d6

In any case, my course folder has the following structure:  

``` bash
.
└── Course4_MoveQ  #(root folder)
    ├─── data-raw  #(contains the file with the GPS locations)
    ├─── docs
    ├─── output    #(contains the cleaned files and processed data)
    ├─── plots
    ├─── R         #(contains the R-script)
    └ Course4_MoveQ.Rproj #(the RStudio-Project file location)
```

## Necessary packages to install and load

We first have to install the packages and load them before we can use the
functions we need.

```{r}
# package names:
pkgs = c("here", "sp", "sf", "dismo", "raster", "GISTools",  "rgdal", 
         "maptools", "rgeos","rgl","rasterVis", "adehabitatHR", "move", "tmap",
         "plotly", "circular","gganimate","moveVis", "ggmap", "maps", "mapproj",
         "viridis","dplyr", "devtools","lubridate", "patchwork","gower",
         "colorspace","ragg", "ggtext","pdftools", "units", "leaflet",
         "glue", "cowplot", "tidyverse", "ggplot2", "Cairo") 

# install.packages(pkgs) # only run this line once for installing the packages!
# update.packages()
```

Tip of the day: If you have already installed some of the packages above, you can
first check which ones are already installed, and save the ones *not installed* in 
an object called 'my_packages' and only install the missing ones:

```{r}
my_packages <- pkgs[!(pkgs %in% installed.packages()[,"Package"])]
my_packages
if(length(my_packages) > 0) install.packages(my_packages)
```

Now load the packages:

```{r, results='hide',echo=FALSE}

# TODO - check if all libraries are needed. However, I guess the problem with 
# course2-distance analysis was that a library was missing for them, and we
# did not realize this because we have so many installed anyway???

```


```{r libraries, warning=FALSE, message=FALSE}
library(here) #for easy directory management
library(sp)
library(dismo)
library(raster)
library(GISTools)
library(rgdal)    # retiring end of 2023 -> use stars/ terra /sf
library(maptools) # retiring end of 2023 -> use sf
library(rgeos)    # retiring end of 2023 -> use sf
library(rgl)
library(rasterVis)
library(adehabitatHR)  #adehabitatLT  #adehabitatMA                           
library(sf)
library(move)
library(plotly)
library(circular)
library(dplyr)
library(tmap)
library(viridis) 
library(gganimate)
library(ggplot2)
library(moveVis)
library(ggmap)
library(maps)  
library(mapproj)
library(devtools)
library(lubridate) 
library(here)
library(glue)
library(cowplot)
library(tidyverse)
library(Cairo)
library(colorspace)
library(ragg)
library(ggtext)
library(pdftools)
library(units)
library(ggplot2)
library(patchwork) # to combine plots
library(leaflet)
library(gower)
```

## Set the working environment

Now that we are working inside an R-project, we can use the easy functionality
of the `{here}` package, which is automatically pointing to your project's root
folder, e.g.:

```{r}
here::here() 
```

Hence, there is no need to use the function *setwd()* any more.
Note: if it does not work, please close RStudio, go to your Explorer and 
double-click on the .Rproj file. Then, under 'files' (usually lower right panel) 
double-click on the R folder and open the script.

## Load data

The movement data are stored in the data-raw subfolder. Let's check which
files are available. 

```{r}
lf <- list.files(path=here("data-raw"), full.names=TRUE) 
lf
```

The output lists two results, there is another subfolder, and many files. E.g.,
the element 7 of the vector lf, lf[7], is a folder. If
you already know that your movement data contains e.g. 'gps' in its name or
is stored as '.txt' or '.csv' files, you can directly search for those files with the
*pattern* argument:

```{r}
# check the difference, and note: full.names is set to FALSE
thefile <- list.files(path=here("data-raw"),pattern='gps',full.names=FALSE)
thefile

# ...and here to TRUE
thefullfile <- list.files(path=here("data-raw"),pattern='gps',full.names=TRUE)
thefullfile
```

Now load the data file. Our first fox filename should be 'tag5334_gps.txt'.
```{r}
# let's get the number of the list in lf that corresponds with the filename.
# Tip of the day: always code everything you can, do not insert numbers or filenames
# by hand, i.e. lf[2]. If you for example add new data, the numbering changes!

lf_number <- which(lf == thefullfile)
dat_anim <- read.table(file=lf[lf_number], header=TRUE, fill=TRUE, sep=',')

# or, since with 'thefullfile' you gave the complete path, you can also directly load it
#dat_anim <- read.table(file=thefullfile, header=TRUE, fill=TRUE, sep=',')

```

# Data check and cleaning

## Checking for missing or incomplete information

Let's have a look at the data. This is the typical way data are stored on e-obs collars:
```{r}
dat_anim[1:5,] # recap: head(dat_anim) also works
```
For now, we will only work with few columns. **tag.serial.number** here refers to the 
individual identifier (collar ID). There are two columns with timestamps. 
**start.timestamp** is the preprogrammed time-interval. Then there is the
**timestamp.of.fix**, which is the real time the GPS-location was recorded. 
This is usually a bit later, i.e. 
<br>
( = **start.timestamp** + **used.time.to.get.fix** + 1 second), as it takes some
time for the collar unit to connect to the satellite. The spatial info is stored
in the columns **longitude** and **latitude**.

Before you can transform the *data.frame* 'dat_anim' into a georeferenced 
spatial object, you need to check whether there are missing locations in your *data.frame*,
otherwise you will get an error message on transformation.
**This can happen if no GPS-signal could be recorded. Or - importantly - some collars
are only activated when the animal is moving to save battery life. In that case,
the missing GPS coordinates would correspond with the last position (= be the same).
Depending on which analysis you want to do, e.g. define resting places, you might need to fill
the missing positions again.**

```{r}
# if the latitude-entry is missing, the longitude value will also be missing
# so it is enough to only check the latitude
which(is.na(dat_anim$latitude)) # there are a lot of missing values in the locations

# delete rows with missing spatial info 
dat_anim_na <- dat_anim[!is.na(dat_anim$latitude),] #alternatively, use complete.cases()
```

Make the crosscheck if there is missing info in a row in longitude. This should NOT be
the case after we had deleted those rows:
```{r}
which(is.na(dat_anim_na$longitude)) # none
```

## Checking for coarse spatial outliers

It could happen the collar was tested e.g. in Berlin, but the animal was finally
caught and collared far away. Make a quick check whether there are strange locations:
```{r}
plot(dat_anim_na$latitude ~ dat_anim_na$longitude)
```
There do not seem to be coarse outliers, data look compact. 


## Appending sunrise and sunset as columns (working with dates)

We will now add some additional columns, where separate days (numbered from 1 to
365), the month (from 1 to 12) and the hour of the day are stored (1 - 23). Sunset and sunrise can be calculated based on dates and the location (latitude, longitude). This can be 
done with the package ´{lubridate}´. Check the [vignette](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html)!

```{r}
# have a look at the timestamp format, here the first row of the column start.timestamp
dat_anim_na$start.timestamp[1]

# define date-time format - the format is year-month-day_hour:min:sec:
dat_anim_na$start.timestamp <- ymd_hms(dat_anim_na$start.timestamp, tz="Europe/Berlin") 
dat_anim_na$start.timestamp[1]
```

Now append the information:

```{r}
dat_anim_na$yearday <- yday(dat_anim_na$start.timestamp)
dat_anim_na$month   <- month(dat_anim_na$start.timestamp)
dat_anim_na$hour    <- hour(dat_anim_na$start.timestamp)
dat_anim_na$kweek   <- week(dat_anim_na$start.timestamp)
dat_anim_na$date    <- date(dat_anim_na$start.timestamp)
# crosscheck with
# head(dat_anim_na)
```

In addition, we want to calculate the hours of sunset and sunrise as well as daylength.
For this, we need to install a package that is still under development, i.e.
which is not on CRAN. We therefore must download and install it locally:

```{r}
# devtools::install_github("bgctw/solartime") # run only once
library(solartime)
```
For computing sunset and sunrise, the latitude/ longitude must be provided as well:
```{r}
dat_anim_na$sunrise   <- computeSunriseHour(timestamp = dat_anim_na$start.timestamp,
                                            latDeg = dat_anim_na$latitude,
                                            longDeg = dat_anim_na$longitude)
dat_anim_na$sunset    <- computeSunsetHour(dat_anim_na$start.timestamp,
                                           dat_anim_na$latitude,
                                           dat_anim_na$longitude)
dat_anim_na$daylength <- computeDayLength(dat_anim_na$start.timestamp,dat_anim_na$latitude)
dat_anim_na$daytime   <- computeIsDayByLocation(dat_anim_na$start.timestamp,
                                                dat_anim_na$latitude,
                                                dat_anim_na$longitude)
# head(dat_anim_na)
```


## Check for temporal outliers

Check if there are strange dates, or dates before you collared the animal (e.g., 
usually the collar is activated and tested before the animal is collared).
Delete these data rows:

```{r}
table(dat_anim_na$date)  #  there is a strange date - 2025-12-26

# Delete data row
delme <- which(dat_anim_na$date == '2025-12-26')
dat_anim_na[delme,]
dat_anim_na <- dat_anim_na[-delme,] # delete the strange date and 
table(dat_anim_na$date)             # check again
plot(table(dat_anim_na$date))       # plot the number of fixes per day
```


## Save the cleaned file

Finally, we save the processed data file that we will use for exploration
and analysis into the subfolder ...'output/data-proc'. There are two options

```{r}
# R data file - can only be opened/ read with R by using function readRDS()
saveRDS(dat_anim_na, file = here('output/data-proc/tag5334_gps_proc.Rds'))

# interchange file format .csv
write.csv(dat_anim_na, file = here('output/data-proc/tag5334_gps_proc.csv'))
```

Check your output folder for these files. The efficient '.Rds' file storage
is about 3 times smaller than the '.csv' file. 


# Exercise 4.1

This is recap from Course2. Please plot the animal relocations in two different 
colours for column 'daytime', using one of 
the options with ´{ggplot2}´,´{ggpmap}´ or ´{leaflet}´. Use the
processed file of fox Q (tag5334_gps_proc) and code it in a separate script. Save 
your script as Course4_Exercise1_*yourname*.R  .
<br><br>
Hint: Remember to load the libraries.
Note that you might want to plot the locations in the correct
spatial dimensions by projecting it using the functions
*st_as_sf()* and *st_transform()* .  

```{r}
#---------------------------------------------------------------------------------------#
```

# Data exploration

## Load the cleaned data file

I recommend that you store the raw file safely and continue with
the cleaned and processed file after major data manipulations have been 
conducted to minimize errors. I'd even suggest to make these steps
in different R-scripts.

```{r}
anim_proc <- readRDS(file = here('output/data-proc/tag5334_gps_proc.Rds'))
# head(anim_proc)
```

Have a look whether there are gaps/ missing days in the data, or whether we have
approximately a regular number of fixes each day.


```{r}
plot(table(anim_proc$date))
```

```{r}
plot(table(anim_proc$yearday)) # what is the difference?
```
The number of fixes (locations taken) per day seems to be quite regular. Mind the 
difference when plotting per date (sorted) and yearday (1-365).

## Plot activity

Now we can have a look at the distribution of the fixes during the day, which is
a hint on the active phase of the animal:

```{r}
plot(table(anim_proc$hour))       # plot the number of fixes per hour
```

More fixes during the night - this is the active phase. As these are circular 
data, we can plot the number of fixes as sign for activity (knowing that our
collars did not record when foxes were inactive! Otherwise, we cannot distinguish
missing data from inactive times! -> bias). 

```{r}
# simple circular plot (rose diagram)
timetoplot <- circular(anim_proc$hour%%24, # convert to 24 hrs = bins
                       units="hours", template="clock24")
rose.diag(timetoplot, bin=24, col="blue",
          main="Events by Hour (sqrt scale)", prop=3)
```

Check whether the activity during the day was before sunrise (our 'daytime' 
column which is either TRUE (=yes, daylight) or FALSE (dark hours):

```{r}
# with ggplot
# code adapted from https://gist.github.com/mattbaggott/4361381
ggplot(anim_proc,aes(x=hour,fill=daytime))+
  geom_bar(breaks=seq(0,24),width = 2,colour="grey")+
  coord_polar(start=0)+
  theme_minimal()+
  scale_fill_brewer()+
  ylab("Count")+
  ggtitle("Events by Time of day")+
  scale_x_continuous("", limits=c(0,24),
                     breaks=seq(0,24),
                     labels=seq(0,24))
```

Apart from the outliers between 11 and 13 o'clock, the active phase was
during the dark hours.

```{r,echo=FALSE, eval=FALSE}
# TODO for later,  not urgent 
# heatmap /activity plot for weeks~hrs (makes only sense for 
# yearly data)
```

## Convert the data into a spatial object

Now we transform the data into *sf* and *sp* objects and **assign** 
the coordinate reference system to :
```{r}
# transform into spatial simple feature sf object
mydf_sf <- st_as_sf(x = data.frame(anim_proc),
                       coords = c("longitude", "latitude"),
                       crs = 4326,
                       sf_column_name = "geometry" )

# transform into SpatialPointsDataFrame  - for crosschecking
mydf_sp <- as(mydf_sf, "Spatial") 
```

And then we **project** the reference system from angular units to a planar
coordinate reference system in meters:

```{r}
# transform CRS to projected one in meter distance units
mydf_sf_trans <-  st_transform(mydf_sf, 3035 )  # EPSG-code  
mydf_sp_trans <-  spTransform(mydf_sp, CRS("+init=epsg:3035")) 
```
Recently, there are issues with the missing datum in the CRS-specifications. We
ignore this for now. More on the issue of moving from proj4 to proj6 in the future: <br>
* https://inbo.github.io/tutorials/tutorials/spatial_crs_coding/
<br>
* https://cran.r-project.org/web/packages/sp/vignettes/CRS_warnings.html
<br>

Make a quick plot of the data:
Recap for styling the plot:<br>
https://www.r-bloggers.com/2021/12/introduction-to-geospatial-visualization-with-the-tmap-package/

```{r}
tmap_mode(mode = "view")
  tm_shape(shp = mydf_sf_trans)  + tm_dots(size = 0.01, 
                                           col = 'daytime',  
                                           alpha=0.5)
```

Have a deep look at the data: did the fox really swim? (points in Lake Rummelsburg).
Or are these outliers? No: the lake was frozen - check the date! Note: you really need to know your animals and the area before analysing data!
<br>

```{r,echo=FALSE, eval=FALSE}
# TODO later - how to access date of the outlier and check temperature ?
# interactive? with labelling?
```

# Basic movement metrics
## Transform your data into a `{move}`object

check here for the possibilities of what to do with the package: 
<br>
* https://cran.r-project.org/web/packages/move/vignettes/move.html <br>
* https://rdrr.io/cran/move/f/vignettes/move.Rmd <br>


```{r, echo=FALSE, results='hide',eval=FALSE}
####### TODO use transformed animal data for spatial measures! #####
####### WTF funktioniert das nicht???????   ########
# Ich wollte andere Koordinaten nehmen, weil unten hrBootstrap 
# komische measures ausspuckt - viel zu kleine HRs

mydf_sf_trans$coords <- st_coordinates(mydf_sf_trans)
tt <- st_drop_geometry(mydf_sf_trans)

my_fox <- move(x = tt$coords.X, y = tt$coords.Y,
               time=as.POSIXct(tt$start.timestamp,format="%Y-%m-%d %H:%M:%S"),
               proj= CRS("+init=epsg:3035"),
               data=tt,
               animal="FoxQvonStralau")  
```


```{r}
my_fox <- move(x = anim_proc$longitude, y = anim_proc$latitude,
               time=as.POSIXct(anim_proc$start.timestamp,format="%Y-%m-%d %H:%M:%S"),
               proj= CRS("+init=epsg:4326"),
               data=anim_proc,
               animal="FoxQvonStralau")  
# head(my_fox); nrow(my_fox)

```

## Time lag, turning angle and speed

### Time lags between the fixes:
```{r}
timeLag(my_fox, unit = "mins")[1:5] # varies a lot, but minimum here 20 min
min(timeLag(my_fox, units="mins"))
mean(timeLag(my_fox, units="mins"))
max(timeLag(my_fox, units="mins"))
```

### Turning angles (makes only sense with high resolution data):
```{r}
turnang <- angle(my_fox)
# turnang <- turnAngleGc(my_fox) # the same
# using the absolute abs(), because -180 and 180 has similar meaning:
# the animal keeps the direction
hist(abs(turnang))
```

### Speed and step length
```{r}
steplength <- distance(my_fox) # know your units!
hist(steplength)
max(steplength) # max 1 km in 20 min = 3 km / h

hist(speed(my_fox)) # units? -> check ?speed
```

# The homerange concept - MCP, kernel, aKDE


See lecture!


### Minimum Convex Polygon

Unfortunately, we have to use *SpatialDataFrame*-Objects for activity range 
calclulation, as the package '{adehabitatHR}' is based on the old '{sp}'-formaat.

```{r, echo=FALSE, results='hide', eval=FALSE}
# TODO - something is wrong here with the units!!!! check the crs warning online
hrBootstrap(my_fox,rep=5, level=99, unout="m2", plot=TRUE) #it is ha, not m2
```

You can calculate the total area covered, or, if you specify a column, you
can calculate the activity range per daytime, animalID or months... Use the 
SpatialDataFrame-Object from the '{sp}' package we created above.

```{r}
mcp_daytime <- mcp(mydf_sp_trans[,'daytime'], percent=95, unout='km2') # MCP (95%) 
mcp_daytime # area for both (daytime separated) polygons

hrs <- mcp.area(mydf_sp_trans[,'daytime'], percent=seq(50, 100, by = 5),unout='km2')
hrs # home range size in km2 for the different MCP-levels

plot(mcp_daytime)
plot(mydf_sp_trans, add=TRUE, col= as.numeric(mydf_sp_trans$daytime)+1)
```
The daytime activity (or better: hiding and resting) range is ten times smaller 
than the nighttime activity.

```{r}
mcp_month <- mcp(mydf_sp_trans[,'month'], percent=95, unout='km2') # MCP (95%) 
mcp_month
```
Note that there is an activity range calculated for October (month 10), but the animal was only collared  30.10.! This definitely does not make sense.
<br>
Now save the MCPs as ESRI shapefiles. 
```{r}
writeOGR(mcp_daytime, dsn=here('output','data-proc'), 
         "mcp95_daytime_foxQ",'ESRI Shapefile',overwrite=T) # save as ESRI shapefile
```

### Kernel utility density

Calculate the density kernel.
```{r}
kud <- kernelUD(mydf_sp_trans[,'daytime'], h="href") # calculate kernel with h="href"
# the output is an object with lists, accessed via: str(kud)
kernel.area(kud, unout='km2') # across all levels 5-95%

kud90 <- getverticeshr(kud,percent = 90) # this creates the spatial object, units m2 
gArea(kud90,byid=T)/1e6                  # units km2
```

Make a plot and save it as spatial polygon. 
```{r}
# plot only the nighttime
image(kud[[1]], col= viridis(100, direction = -1))
xyz <- as.image.SpatialGridDataFrame(kud[[1]])
contour(xyz, add=TRUE)
points(mydf_sp_trans, cex= 0.01, col='red')
```


```{r}
# save it as shapefile - mind - here we can only save one contour line, 
# in this case the 90% kernel:
writeOGR(kud90, dsn=here('output','data-proc'), 
         layer="kernel_ud90",'ESRI Shapefile',overwrite=T)
```


```{r}
# make the crosscheck: load and plot
kernel_ud90_sf <- st_read(dsn=here('output','data-proc'),layer="kernel_ud90")
mcp95_sf       <- st_read(dsn=here('output','data-proc'),layer="mcp95_daytime_foxQ")
# note: the CRS is lost and not defined! So assign it again:
st_crs(kernel_ud90_sf) <- 3035
st_crs(mcp95_sf)       <- 3035
```


```{r}
# TODO eigentlich wollte ich alles übereinander plotten, also mcp, kernel, punkte

ggplot(kernel_ud90_sf) +
    geom_sf() 
 #  + geom_sf(mydf_sf_trans, size = 1, shape = 23, fill = "darkred")
```


Do the following only on a fast computer:
```{r,echo=TRUE}
##### der krass coole 3D plot

#install.packages("devtools")
#devtools::install_github("ropensci/plotly")
#library(plotly)
my_kud <- kud[[1]] # or choose kud[[2]]

xy <- coordinates(my_kud)
z <- my_kud@data$ud
df_kud <- data.frame(x=xy[,1],y=xy[,2],z=z)
#persp(x=df_kud$x, y=df_kud$y, z= df_kud$z)
#plot3d(x=xy[,1],y=xy[,2],z=z)

my_kud@grid@cells.dim
r <- raster(ncols=60,nrows=58)
coordinates(df_kud) <- ~x + y
r_kud <- rasterize(df_kud,r,'z',fun=max)

myz <- matrix(z ,nrow= my_kud@grid@cells.dim[[2]],
              ncol= my_kud@grid@cells.dim[[1]],byrow=F)
# image(myz)
kd.list <- list(x=xy[,1],y=xy[,2],z=myz)
# with(kd.list, plot_ly(x = x, y = y, z = z, type = "surface"))
plot_ly(z = myz, type = "surface") # click into plot and on icon 'turntable rotation' in plot upper right
#########  ende krass cooler plot ########
```


# Nothing makes sense but in the light of environmental information

-> Exercise 4.2

# Exercise 4.2





```{r, eval=FALSE, echo=FALSE, results='hide'}

# TODO ignore this for now


# TODO # something I really don't understand are the different sizes
# in mcp kernel ctmm - > the units are wrong, it is either ha, m2 or km2, but
# it is never correct in all three applications


#############################################################################
######## advanced home range analyses - variogram and  aKDE  ################
#############################################################################

library(ctmm)
# please read the vignettes of ctmm package to understand what is done in the following
# https://cran.r-project.org/web/packages/ctmm/vignettes/variogram.html
# https://cran.r-project.org/web/packages/ctmm/vignettes/akde.html

# This package needs a 'telemetry' object, which can either be created from a 
# 'data.frame' or can use our move-Object from above.

 # anim_proc$individual.local.identifier <- anim_proc$tag.serial.number
 # anim_proc$timestamp                   <- anim_proc$timestamp.of.fix
 # anim_proc$location.long               <- anim_proc$longitude
 # anim_proc$location.lat                <- anim_proc$latitude
 # ctmm_anim <- as.telemetry(anim_proc)

ctmm_anim <- as.telemetry(my_fox)


ctmm_anim <- as.telemetry(my_fox[1:1000,]) #I am restricting the data to the first points for saving computation time
plot(ctmm_anim) #how strange! seems to be the DOP
# https://cran.r-project.org/web/packages/ctmm/vignettes/error.html
help(plot.telemetry)
ctmm::plot(ctmm_anim, pch='.',lwd=0.1,cex=0.00001,xlim= c(-2000,2000)) #something does not work here....

# please do the following steps for fitting the best model (movement process) to your data
# https://cran.r-project.org/web/packages/ctmm/vignettes/variogram.html

SVF <- variogram(ctmm_anim)
plot(SVF,fraction=0.10) # TODO # #Achtung, hectares ! something is wrong with the units, should be 0.55 km2 or 55 ha!
title("zoomed in")
# this info gives first parameters for sigma, tau etc for fitting m.ouf (see below)
# restricted space use (plateau),
# autocorrelated positions ()
# and autocorrelated velocities (upward curvature)

plot(SVF,fraction=0.55,level=c(0.5,0.95))
#plot(SVF,fraction=1,level=c(0.5,0.95))
title("zoomed out")


m.ouf <- ctmm(sigma=6 %#% "hectares",tau=c(0.04 %#% "day",0.2 %#% "hour"))
plot(SVF,CTMM=m.ouf,level=level,col.CTMM="blue",fraction=0.02) #does the blue line fit?
title("Ornstein-Uhlenbeck-F movement")

### please read the variogram-vignette for 'Irregular Sampling Schedules'
# if your data was not sampled in regular intervals
# 20, 40 min sampling intervals
level <- c(0.5,0.95) # 50% and 95% CIs
dt <- c(20,40) %#% "min"
SVF3 <- variogram(ctmm_anim,dt=dt)
plot(SVF3,fraction=0.55,level=level)
title("Multi method")

# students ignore this!
#scale_length_ov_vector <- round(length(SVF3@.Data[[1]])*0.8,digits=0) - 1000
#plot(SVF3@.Data[[1]][25:scale_length_ov_vector],type='l') #for fitting colour of noise
#forcolor <- SVF3@.Data[[1]]
#write.csv(forcolor, 'Q_forCol.csv')


### Maximum likelihood fitting
### for a first 'guessimate' of model parameters
## not run! - takes a long time for many data points
#GUESS <- ctmm.guess(ctmm_anim,interactive=FALSE)
#FIT <- ctmm.fit(ctmm_anim,GUESS)
#summary(FIT)


########################   aKDE fit ####################
# https://cran.r-project.org/web/packages/ctmm/vignettes/akde.html

M.IID <- ctmm.fit(ctmm_anim) # no autocorrelation timescales - yields KUD
summary(M.IID)

M.OUF <- ctmm.fit(ctmm_anim,m.ouf) # m.ouf quickly fitted by hand - see above
summary(M.OUF)

m.ouf <- ctmm.guess(ctmm_anim,interactive=FALSE) # automated model guess
M.OUF <- ctmm.fit(ctmm_anim,m.ouf)
summary(M.OUF)

UD0 <- akde(ctmm_anim,M.IID)
UD2 <- akde(ctmm_anim,M.OUF)
UD2w <- akde(ctmm_anim,M.OUF,weights=TRUE) # for irregular sampling intervals
summary(UD0);summary(UD2);summary(UD2w)

# calculate one extent for all UDs
EXT <- extent(list(UD0,UD2,UD2w),level=0.95)
EXT

par(mfrow=c(2,2))
plot(ctmm_anim,UD=UD0,xlim=EXT$x,ylim=EXT$y)
title(expression("IID KDE"["C"]))
plot(ctmm_anim,UD=UD2,xlim=EXT$x,ylim=EXT$y)
title(expression("OUF AKDE"["C"]))
plot(ctmm_anim,UD=UD2w,xlim=EXT$x,ylim=EXT$y)
title(expression("weighted OUF AKDE"["C"]))



############################################################################################
library(amt)
#continue with SSFs for selection analysis
#https://cran.r-project.org/web/packages/amt/vignettes/p4_SSF.html

############################################################################################
########## animate movement path with moveVis - take care, takes a lot of computation time !
############################################################################################

### not run!
### aprox 30 min for 300 frames (in my computer)
# http://movevis.org/index.html#get-started
#move_data <- align_move(my_fox, res = 20, unit = "mins")
#frames <- frames_spatial(move_data[1:100,], # restricted to only 100 data points!
#                         path_colours = c("blue"),
#                         map_service = "osm",
#                         map_type = "topographic",
#                         alpha = 0.5)
## alternative with satellite data as background
##frames <- frames_spatial(move_data[1:100,], map_service = "mapbox", map_type = "satellite",
##                         map_token = "YOUR_MAPBOX_TOKEN")
#length(frames)
#animate_frames(frames, out_file = "example_1.gif",overwrite=TRUE)

```

```{r, echo=FALSE, eval=FALSE, results='hide'}
TODO 

- work with d6berlin - load imperviousness and create random points and check the difference
- recurse analysis - favourite places of Q
```


# END

Adding the session info can be very helpful when going back to old scripts or 
using scripts of others:

<details><summary>Session Info</summary>

```{r session-info}
sessionInfo()
```

</details>

